---
title: "Chapter 7 Homework"
output: html_notebook
---

```{r}
library(mosaic)
require(mosaic)
require(mosaicData)
```

## SAT Exploration

In the SAT data `SAT.csv` , the variables have these units:
• sat has units of "points."
• expend has units of "dollars."
• ratio has units of "students."
• frac has units of "percentage points."

Find the model formula for `sat ~ expend + frac`
```{r}
mod = lm(sat ~ expend + frac, data = SAT)
```

What are the coefficients of the model?
```{r}
coef(mod)
```

What is the model explaining and what are the variables used to try to explain
that data?

The model is sat score explained by expend and frac


Consider the model formula `coef(mod)`

(a) What are the units of the coefficient 994?
    A points
    B dollars
    C students
    D percentage points
    E points per dollar
    F students per point
    G points per student
    H points per percentage points
    
    points
    
(b) What are the units of the coefficient 12.29?
    A points
    B dollars
    C students
    D dollars per student
    E points per dollar
    F students per point
    G points per student
    
    points per dollar
    
(c) What are the units of the coefficient 2.85?
    A points
    B dollars
    C percentage points
    D points per dollar
    E students per point
    F points per student
    G points per percentage points
    
    points per percentage point


## More SAT Modeling

Find the model formulas explaining `math` SAT scores using first `salary`, then `frac`, then both, and then include an interaction term.

What do your more sophisticated models tell you that you couldn't see in the simple models?
```{r}
Rail = RailTrail
```


## Rail Trail modeling
`RailTrail` is a collection of data of trail use for 90 days. It is A data frame with 90 observations on the following variables.

hightemp daily high temperature (in degrees Fahrenheit)
lowtemp daily low temperature (in degrees Fahrenheit)
avgtemp average of daily low and daily high temperature (in degrees Fahrenheit)
spring indicator of whether the season was Spring
summer indicator of whether the season was Summer
fall indicator of whether the season was Fall
cloudcover measure of cloud cover (in oktas)
precip measure of precipitation (in inches)
volume estimated number of trail users that day (number of breaks recorded)
weekday logical indicator of whether the day was a non-holiday weekday
dayType one of "weekday" or "weekend"

What explanatory variables would you use to explain the variation of the response variable `volume`? 

weekday and precip

Create 4 models to explain the variation of the number of rail trail users. 
 * A model with one categorical explanatory variable
 * A model with one numeric explanatory variable
 * A model with both a categorical and a numeric explanatory variable
 * A model that you think best explains the data.

```{r}
mod1 = lm(volume ~ weekday, data = Rail)
mod2 = lm(volume ~ avgtemp, data = Rail)
mod3 = lm(volume ~ fall + cloudcover, data = Rail)
mod4 = lm(volume ~ avgtemp + weekday, data = Rail)
summary(mod1)
summary(mod2)
summary(mod3)
summary(mod4)
```

