---
title: "09-Partitioning"
output: html_notebook
---

# Correlation and partitioning of variation

```{r include = FALSE}
library(mosaic)
require(mosaic)
require(mosaicData)
options(width = 100)
```

The coefficient of determination, $R^2$, 
compares the variation in the response variable to the variation
in the fitted model value.  It can be calculated as a ratio of variances, that is the fraction of the total variance that can be explained by the model: 

```{r}
Swim <- SwimRecords # from mosaicData
mod <- lm( time ~ year + sex, data = Swim)
var(fitted(mod)) / var(Swim$time)
```

The convenience function `rsquared()` does the calculation for you:
```{r}
rsquared(mod)
```

The regression report is a standard way of summarizing
models.  Such a report is produced by most statistical software
packages and used in many fields.  The first part of the table
contains the coefficients --- labeled "Estimate" --- along with
other information that will be introduced starting in Chapter
\@ref("chap:confidence").  The $R^2$ statistic is a standard part of the
report; look at the second line from the bottom.
```{r}
summary(mod)
```


Occasionally, you may be interested in the correlation coefficient $r$ between two
quantities.  
You can, of course, compute $r$ by fitting a model, finding $R^2$,
and taking a square root.
```{r}
mod2 <- lm( time ~ year, data = Swim)
coef(mod2)
sqrt(rsquared(mod2))
```

The `cor()` function computes this directly:
```{r}
cor(Swim$time, Swim$year)
```

Note that the negative sign on $r$ indicates that record swim `time`
decreases as `year` increases.  This information about the
direction of change is contained in the sign of the coefficient from
the model.  The magnitude of the coefficient tells how fast the
`time` is changing (with units of seconds per year).  The
correlation coefficient (like $R^2$) is without units.

Keep in mind that the correlation coefficient $r$ summarizes only the
simple linear model A ~ B where B is quantitative.  But the
coefficient of determination, $R^2$, summarizes any model; it is much
more useful.  If you want to see the direction of change, look at the
sign of the correlation coefficient.

# Homework

### Thinking Problem 1

For each of the following pairs of models, mark the statement
that is most correct.

### Part 1

Model 1 . `A ~ B+C`
Model 2 . `A ~ B*C`

  A Model 1 is nested in Model 2.
  
  B Model 2 is nested in Model 1.
  
  C The two models are the same.
  
  D None of the above is true.
  
  A

### Part 2

Model 1 . `A ~ B`
Model 2 . `B ~ A`

  A Model 1 is nested in Model 2.

  B Model 2 is nested in Model 1.

  C The two models are the same.

  D None of the above is true.
  
  D

### Part 3

Model 1 . `A ~ B + C`
Model 2 . `B ~ A * C`

  A Model 1 is nested in Model 2.

  B Model 2 is nested in Model 1.

  C The two models are the same.

  D None of the above is true.
  
  D

### Part 4

Model 1 . `A ~ B + C + B:C`
Model 2 . `A ~ B * C`

  A Model 1 is nested in Model 2.

  B Model 2 is nested in Model 1.

  C The two models are the same.

  D None of the above is true.
  
  D
  
## Thinking Problem 2

For each of the following pairs of models, mark the statement that is most correct.

### Part 1

Model 1 . `A ~ B+C`
Model 2 . `A ~ B*C`

  A Model 1 can have a higher R2 than Model 2

  B Model 2 can have a higher R2 than Model 1

  C The R2 values will be the same.

  D None of the above are necessarily true.
  
  B

### Part 2

Model 1 . `A ~ B + C`
Model 2 . `B ~ A * C`

  A Model 1 can have a higher R2 than Model 2

  B Model 2 can have a higher R2 than Model 1

  C The R2 values will be the same.

  D None of the above are necessarily true.
  
  D

### Part 3

Model 1 . `A ~ B + C + B:C`
Model 2 . `A ~ B * C`

  A Model 1 can have a higher R2 than Model 2

  B Model 2 can have a higher R2 than Model 1

  C The R2 values will be the same.

  D None of the above are necessarily true.
  
  D

[From a suggestion by a student]

## Going further

In answering this question, you might want to try out a few examples using real data: just pick two quantitative variables to stand in for `A` and `B`.
What will be the relationship between R2 for the following two models?

Model 1 . `A ~ B`
Model 2 . `B ~ A`

  A Model 1 can have a higher R2 than Model 2

  B Model 2 can have a higher R2 than Model 1

  C The R2 values will be the same.

  D None of the above are necessarily true.
  
  C

## Application Problem 

Choose one of the models that you have created for an assignment in class. Calculate its \(R^2\) value and, if it is a simple linear model, its correlation coefficient. Write a sentence that explains in the language of the data what \(R^2\) is telling you about the model.
```{r}

```

