---
title: "09-Partitioning"
output: html_notebook
---

# Modeling randomness

```{r include = FALSE}
library(mosaic)
options(width = 100)
```

```{r, echo=FALSE}
set.seed(42)
```
One of the most surprising outcomes of the revolution in computing
technology has been the discovery of diverse uses for randomness 
in the analysis of data and in science generally.  Most young people
have little trouble with the idea of a computer generating random data;
they see it in computer games and simulations.  Older people,
raised with the idea that computers do mathematical operations and
that such operations are completely deterministic, sometimes find 
computer-generated randomness suspect.  Indeed, conventional algebraic
notation ($+$, $-$, $\sqrt{ }$, $\cos$, and so on) has no
notation for "generate at random."

One of the simplest operators for generating random events is
`resample`.  This takes two arguments: the first is a set of
items to choose from at random, the second is how many events
to generate.  Each item is equally likely to be
choosen.  For example, here is a simulation of a coin flip:
```{r}
coin <- c("H","T")
resample(coin, 5)
resample(coin, 5)
```
The first command creates an object holding the possible outcome of
each event, called `coin`.  The next command generated five
events, each event being a random choice of the outcomes in
`coin`.

Another example is rolling dice.  First, construct a set of the
possible outcomes: the numbers 1, 2, 3, 4, 5, 6.
```{r}
die <- seq(1,6)
die
```
Then generate random events.  Here is a roll of two dice.
```{r}
resample(die,2)
```

The `resample()` function is also useful for selecting cases at
random from a data frame.  You have already seen it put to use in generating sampling
distributions via the bootstrap.  This technique will be further developed in later chapters. 


## Random Draws from Probability Models

Although `resample()` is useful for random sampling, it can work
only with finite sets of possible outcomes such as H/T or 1/2/3/4/5/6
or the cases in a data frame.  By default in `resample()`, the 
underlying probability model is equiprobability
 --- each possible outcome is equally likely.
You can specify another probability model by using the `prob = ` argument
to `resample()`. For instance, to flip coins that are very likely (90% of the time, on average) to come up heads:
```{r}
resample( coin, 10, prob = c(.9,.1))
```


R provides other operators that allow
draws to be made from outcome sets that are infinite.
 

For example, the `rnorm()` function makes random draws from a
normal probability distribution.  The required argument tells how many
draws to make.  Optional, named arguments let you specify the mean and
standard deviation of the particular normal distribution that you
want.  To illustrate, here is a set of 15 random numbers from a normal
distribution with mean 1000 and standard deviation 75:
```{r}
samps <- rnorm(15, mean = 1000, sd = 75)
samps
```
In this example, the output was assigned to an object `samps`
to facilitate some additional computations to the
values.  For instance, here is the mean and standard deviation of the
sample:
```{r}
mean(samps)
sd(samps)
```
Don't be surprised that the mean and standard deviation of the sample
don't match exactly the parameters that were set with the arguments
`mean = 1000, sd = 75`.  The sample was drawn at random and so the
sample statistics are going to vary from one sample to the next.  Part
of the statistical methodology to be studied in later chapters has to
do with determining how close the statistics calculated from a sample
are likely to be to the parameters of the underlying population.

Often you will generate very large samples.  In these situations you
usually don't want to display all the samples, just do calculations
with them.  The practical limits of "large" depend on the computer
you are using and how much time you are willing to spend on a
calculation.   For an operator like `rnorm` and the others to be
introduced in this chapter, it's feasible to generate samples of size
10,000 or 100,000 on an ordinary laptop computer.  
```{r}
samps <- rnorm(100000, mean = 1000, sd = 75)
mean( samps )
sd( samps )
```
Notice that the sample mean and standard deviation are quite close to
the population parameters in this large sample.  (Remember not to put commas in as punctuation in large numbers: it's `100000` not 100,000.)

The simulations that you will do in later chapters
will be much more elaborate than the simple draws here.  Even with 
today's computers, you will want to use only a few hundred trials.

## Standard Probability Models

R provides a large set of operators like `rnorm` for different
probability models.  All of these operators work in the same way:
<!-- begin itemize -->

* Each has a required first argument that gives the number of
draws to make.

* Each has an optional set of parameters that specify the
particular probability distribution you want.
<!-- end itemize -->

All the operators start with the letter `r` --- standing for
"random" --- followed by the name of the probability model:


Family | R name |    Parameters    | Nature        
-------|--------|------------------|---------
Normal | `rnorm` | `mean`,`sd` | continuous
Uniform | `runif` | `min`,`max` | continuous
Binomial | `rbinom`| `size`,`prob` | discrete
Poisson | `rpois`| Average rate (written `lambda`)| discrete 
Exponential | `rexp`| Same rate as in poisson but the parameter is called `rate`.| continuous 
Lognormal | `rlnorm`| Mean and sd of the natural logarithm. `meanlog`, `sdlog`| continuous 
$\chi^2$ | `rchisq` | Degrees of freedom (`df`)|continuous 
t | `rt` | Degrees of freedom (`df`)|continuous
F | `rf` | Degrees of freedom in the numerator and in the denominator (`df1`, `df2`)|continuous


To use these operators, you first must choose a particular probability
model based on the setting that applies in your situation.  This
setting will usually indicate what the population parameters should
be.  Some examples:
<!-- begin itemize -->


* You are in charge of a hiring committee 
that is  going to interview three candidates selected from a
population of job applicants that is 63% female.  How many of the
interviewees will be female?  Modeling this as random selection from
the applicant pool, a binomial model is appropriate.  The `size`
of each trial is 3, the probability of being female is 63% :
```{r}
samps <- rbinom(40, size = 3, prob = 0.63)
samps
```
There are 40 trials here, since the first argument was set to 40.
Remember, each of the trials is a simulation of one hiring event.  In
the first simulated event, two of the interviewees were female; in the
third only one was female.  Typically, you will be summarizing all the
simulations, for example to see how likely each possible outcome is.
```{r}
table(samps)
```
* You want to simulate the number of customers who come into a
store over the course of an hour.  The average rate is 15 per hour.
To simulate a situation where customers arrive randomly, the poisson
model is appropriate:
```{r}
rpois(25, lambda = 15)
```
* You want to generate a simulation of the interval between
earthquakes as in Example \@ref("example:earthquake-intervals").  To
simulate the random intervals with a typical rate of 0.03 earthquakes
per year, you would use
```{r}
rexp( 15, rate = 0.03 )
```
Notice the huge variation in the intervals, from less than two months
to more than 210 years between earthquakes.

<!-- end itemize -->

## Quantiles and Coverage Intervals


You will often need to compute coverage intervals in order to describe 
the range of likely outcomes from a random process.  R provides a
series of operators for this purpose; a separate operator for each
named probability model.  The operators all begin with `q`,
standing for quantiles.  In all cases, the first argument is
the set of quantiles you want to calculate for the particular
probability model.  The optional named arguments are the parameters.

Remember
that to find a 95%  coverage interval you need the 0.025 and 0.975
quantiles.  For a 99%  interval, you need the 0.005 and 0.995
quantiles.

To illustrate, here are 95%  coverage intervals for a few
probability models. 
<!-- begin itemize -->

* A normal distribution with mean 0 and standard deviation 1:
```{r}
qnorm( c(0.025, 0.975), mean = 0, sd = 1)
```


* The hiring committee situation modelled by a binomial
distribution with `size = 3` and `prob = 0.63`:

```{r}
qbinom( c(0.025, 0.975), size = 3, prob = 0.63)
```

Perhaps you are surprised to see that the coverage interval includes
all the possible outcomes.  That's because the number of cases in each
trial ($n = 3$) is quite small. 

* The number of customers entering a store during an hour 
as modelled by a poisson
distribution with an average rate of 15 per hour.
```{r}
qpois( c(0.025, 0.975), lambda = 15)
```

* The interval between earthquakes modelled by an exponential
distribution with a typical rate of 0.03 earthquakes per year:
```{r}
qexp( c(.025, .975), rate = 0.03)
```
<!-- end itemize -->


You can also use the `q` operators to find the value that would
be at a particular percentile.  For example, the exponential model with `rate = 0.03` 
gives the 25th percentile of
the interval between earthquakes as:
```{r}
qexp( .25, rate = 0.03)
```
A quarter of the time, the interval between earthquakes will be 9.59
years or less.

It's entirely feasible to calculate percentiles and 
coverage intervals by combining
the random-number generators with `quantile`.  For example, here
is the 95% coverage interval from a normal distribution with mean 0
and standard deviation 1:
```{r}
samps <- rnorm(10000, mean = 0, sd = 1)
qdata( samps, c(.025, .975) )
```

The disadvantage of this approach is that it is a simulation and the
results will vary randomly.  By making the sample size large enough
--- here it is $n = 10000$ --- you can reduce the random variation. 
Using the `q` operators uses mathematical analysis to give 
you what is effectively an
infinite sample size.  For this reason, it's advisable to use the
`q` operators when you can.  However, for many of the techniques to
be introduced in later chapters you will have to generate a random
sample and then apply `quantile` to approximate the coverage intervals.

## Percentiles


A percentile computation applies to situations where you have a
measured value and you want to know where that value ranks relative to
the entire set of possible outcomes.  You have already seen
percentiles computed from samples; they also apply to probability models.

It's easy to confuse percentiles with quantiles because they are so
closely related.
Mathematically, the percentile operators are the inverse of the
quantile operators. 
To help you remember which is which, it's helpful to distinguish them
based on the type of argument that you give to the operator:
<!-- begin description -->

* **Percentile** (e.g. `pnorm`)
The input 
argument is a measured value, something that could be the
output of a single draw from the probability distribution. The output
is always a number between 0 and 1 --- a percentile. 

* **Quantile** (e.g. `qnorm`)
The input is a percentile, a number between 0 and 1.
The output is on the scale of the measured variable.
<!-- end description -->


Example:  You have just gotten your score, 670, on a professional school
admissions test.  According to the information published by the
testing company, the scores are normally distributed with a mean of
600 and a standard deviation of 100.  So, your ranking on the test, as
indicated by a percentile, is:

```{r}
pnorm(670, mean = 600, sd = 100)
```
Your score is at about the 75th percentile.  

Example: Unfortunately, the professional school that you want to go to
accepts only students with scores in the top 15 percent.  Your score,
at 75.8%, isn't good enough.  So, you will study some more and take
practice tests until your score is good enough.
How well will you need to score to reach the 85th percentile?
```{r}
qnorm(0.85, mean = 600, sd = 100)
```

# Homework

### Problem 1

Two basic operations that you need to perform on probability models are these:

**percentile** Given a value, what is the probability (according to the model) of the outcome from one trial being that value or less?

**quantile** Given a probability, what is the value whose percentile is that probability?

To illustrate by example, suppose that you are dealing with a probability model for an IQ test score that is a normal distribution with these parameters: mean = 100 and standard deviation = 15.

Percentile question: What is the percentile that corresponds to a test score of 120? Answer: 0.91 or, in other words, the 91st percentile.

```{r}
pnorm(120, mean=100,sd=15)
```

Quantile question: What score will 95% of scores be less
than or equal to? Answer: a score of 125.

```{r}
qnorm(0.95, mean=100, sd=15)

```

Here are two very basic questions about percentile and quantile calculations:

 (a) True or False The output of a percentile question will always be a probability, that is, a number between 0 and 1.
 
 True
 
 (b) True or False The output of a quantile question will always be a value, that is, something in the same units as the random variable.
 
 True

Sometimes to answer more complicated questions, you need first to answer one or more percentile or quantile questions.

Answer the following questions, using the normal probability model with the parameters given above:

 (a) What’s the test score that almost everybody, say, 99% of people, will do better than?
 
```{r}
qnorm(.01, mean=100,sd=15)
```
 
 • Which kind of calculation is this? quantile

 • What is the answer? 65

 (b) To calculate a coverage interval on a probability model, you need to calculate two quantities: one for the left end of the interval and one for the right. Which type of calculation are these probabilities from:
 
 quantile

  (c) Calculate a 50% coverage interval on the test scores, that is the range from the 0.25 quantile to the 0.75 quantile:
  
```{r}
qnorm(.25, mean=100,sd=15)
qnorm(.75, mean=100,sd=15)
```
  
  • Left end of interval: 90

  • Right end of interval: 110
 
 (d) Calculate an 80% coverage interval, that is the range from the 0.10 to the 0.90 quantile:
 
```{r}
qnorm(.1, mean=100,sd=15)
qnorm(.9, mean=100,sd=15)
```
 
  • Left end of interval: 81
  
  • Right end of interval: 119
  
 (e) To calculate the probability of an outcome falling in a given range, you need to do two percentile calculations, one for each end of the range. Then you subtract the two different probabilities. What is the probability of a test score falling between 100 and 120?
 
```{r}
pnorm(120, mean=100,sd=15)-pnorm(100, mean=100,sd=15)
```
 
 .41
    
### Problem 2

A coverage interval gives a range of values. The “level” of the
interval is the probability that a random trial will fall inside
that range. For example, in a 95% coverage interval, 95% of
the trials will fall within the range.

To construct a coverage interval, you need to translate the
level into two quantiles, one for the left side of the range and
one for the right side. For example, a 50% coverage interval
runs from the 0.25 quantile on the left to the 0.75 quantile on
the right; a 60% coverage interval runs from 0.20 on the left
to 0.80 on the right. The probabilities used in calculating the
quantiles are set so that

  • the difference between them is the level of the interval.
For instance, 0.75 and 0.25 give a 50% interval.

  • they are symmetric. That is, the left probability should
be exactly as far from 0 as the right probability is from 1

A classroom of students was asked to calculate the left
and right probabilities for various coverage intervals. Some of
their answers were wrong. Explain what is wrong, if anything,
for each of these answers.

 (a) For a 70% interval, the 0.20 and 0.90 quantiles 
    
    A The difference between them isn’t 0.70
    B They are not symmetrical.
    C Nothing is wrong.
    
    B
 
 (b) For a 95% interval, the 0.05 and 0.95 quantiles.

    A The difference between them isn’t 0.95
    B They are not symmetrical.
    C Nothing is wrong.
    
    A
 
 (c) For a 95% interval, the 0.025 and 0.975 quantiles.
    
    A The difference between them isn’t 0.95
    B They are not symmetrical.
    C Nothing is wrong.
    
    C

### Problem 3

For each of the following probability models, calculate a 95%
coverage interval. This means that you should specify a left
value and a right value. The left value corresponds to a probability of 0.025 and the right value to a probability of 0.975.

  (a) The number of cars driving along a highway in one hour, when the mean number of cars is 2000 per hour. Hint: Poisson model
  
```{r}
qpois( c(0.025, 0.975), lambda = 2000)
```
  
    Left side of interval: 1913
    Right side of interval: 2088

  (b) The number of heads out of 100 flips of a fair coin. Hint:
Binomial model.

```{r}
qbinom(c(0.025, 0.975), size = 100, prob = 0.5)
```

    Left side of interval: 40
    Right side of interval: 60
    
  (c) The angle of a random spinner, ranging from 0 to 360 degrees. Hint: Uniform model.
  
```{r}
qunif(c(0.025, 0.975), 0, 360)
```
  
  
    Left side of interval: 9
    Right side of interval: 351


### Problem 4

College admissions offices collect information about each
year’s applicants, admitted students, and matriculated students. At one college, the admissions office knows from past
years that 30% of admitted students will matriculate.

The admissions office explains to the administration each
year that the results of the admissions process vary from year
to year due to random sampling fluctuations. Each year’s
results can be interpreted as a draw from a random process
with a particular distribution.

Which family of probability distribution can best be used
to model each of the following situations?

  (a) 1500 students are offered admission. The number of students who will actually matriculate is:

    A Normal
    B Uniform
    C Binomial
    D Poisson
    E Exponential
    F Lognormal
    
    C

  (b) The average SAT score of the admitted applicants:
  
    A Normal
    B Uniform
    C Binomial
    D Poisson
    E Exponential
    F Lognormal
    
    A

  (c) The number of women in the matriculated class:

    A Normal
    B Uniform
    C Binomial
    D Poisson
    E Exponential
    F Lognormal
    
    A

### Problem 5

For each of the following, decide whether the random variable
is binomial or not. Then choose the best answer from the set
offered.

  (a) Number of aces in a draw of 10 cards from a shuffled deck
with replacement.

    A It is binomial.
    B It’s not because the sample size is not fixed.
    C It’s not because the probability is not fixed
    for every individual component.
    D It’s not for both of the above reasons.
    
    A
    
  (b) Number of aces in a draw of 10 cards from a shuffled deck
without replacement.

    A It is binomial.
    B It’s not because the sample size is not fixed.
    C It’s not because the probability is not fixed
    for every individual component.
    D It’s not for both of the above reasons.
    
    C

  (c) A broken typing machine has probability of 0.05 to make
a mistake on each character. The number of erroneous
characters in each sentence of a report.

    A It is binomial.
    B It’s not because the sample size is not fixed.
    C It’s not because the probability is not fixed
    for every individual component.
    D It’s not for both of the above reasons.
    
    B

  (d) Suppose screws produced by a certain company will be defective with probability 0.01 independent of each other.
The company sells the screws in a package of 10. The
number of defective screws in a randomly selected pack.

    A It is binomial.
    B It’s not because the sample size is not fixed.
    C It’s not because the probability is not fixed for every individual component.
    D It’s not for both of the above reasons.
    
    A
    
  (e) Observe the sex of the next 50 children born at a local
hospital. Let x= # of girls among them.

    A It is binomial.
    B It’s not because the sample size is not fixed.
    C It’s not because the probability is not fixed
    for every individual component.
    D It’s not for both of the above reasons.
    
    A

  (f) A couple decides to continue to have children until their
first daughter. Let x = # of children the couple has.

    A It is binomial.
    B It’s not because the sample size is not fixed.
    C It’s not because the probability is not fixed
    for every individual component.
    D It’s not for both of the above reasons.
    
    B

  (g) Jason buys the state lottery ticket every month using his
favorite combination based on his birthday and his wife’s.
x= # of times he wins a prize in one year.

    A It is binomial.
    B It’s not because the sample size is not fixed.
    C It’s not because the probability is not fixed for every individual component.
    D It’s not for both of the above reasons.
    
    A

