---
title: "CS205P2"
author: "Andrew Poock"
date: "2024-02-12"
output: html_document
---


```{r, include=FALSE}
library(mosaic)
library(dcData)
library(tidyverse)
OB <- OrdwayBirds %>%
  select (SpeciesName, Month, Day) %>%
  mutate(Month = as.numeric(as.character(Month)),
  Day = as.numeric(as.character(Day)))
```

#### Question 1
```{r}
n_distinct(OB$SpeciesName)
```

#### Question 2

Variables used to match cases: SpeciesName
```{r, warning=FALSE}
Corrected <- OrdwayBirds %>%
  inner_join(OrdwaySpeciesNames) %>%
  select(Species=SpeciesNameCleaned, Month, Day) %>%
  na.omit()
```

#### Question 3

Variables added: Species / SpeciesNameCleaned

#### Question 4
```{r}
Captures <- Corrected %>%
  group_by(Species) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
head(Captures, 10)
```

Graph Setup
```{r}
# Major Species: n > 1000
Majors <- Corrected %>%
  filter(Species %in% c("Slate-colored Junco", "Tree Swallow", "Black-capped Chickadee", "American Goldfinch", "Field Sparrow"))
ByMonth <- Majors %>%
  group_by(Species, Month) %>%
  summarize(count = n())

gf_bar(~ Month, data = ByMonth, color = "black", fill = ~ Species) |>
  gf_labs(title = "Major Bird Species By Month", x = "Month", y = "Number of Species") |> 
  gf_theme(legend.position = "right")

print(ByMonth)
```

#### Question 5

American Goldfinch and Black-capped Chickadee are present year round

#### Question 6

Field Sparrow, Slate-colored Junco, Tree Swallow are migratory

#### Question 7

Peak Months

* American Goldfinch: October
* Black-capped Chickadee: November
* Field Sparrow: May
* Slate-colored Junco: October
* Tree Swallow: March

#### Question 8
```{r}
medians <- ByMonth %>%
  group_by(Species) %>%
  summarize(median = median(count))
head(medians, 5)
```

```{r}
good <- ByMonth %>%
  left_join(medians) %>%
  filter(count > median) %>%
  group_by(Species) %>%
  summarize(NMonths = n()) %>%
  filter(NMonths >= 6)
head(good)
```

